---
title: "Darwin Core mapping"
subtitle: "For: Inventory of alien macroinvertebrates in Flanders, Belgium"
author:
- Lien Reyserhove
- Peter Desmet
- Dimitri Brosens
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
---

This document describes how we map the checklist data to Darwin Core. The source file for this document can be found [here](https://github.com/trias-project/alien-fishes-checklist/blob/master/src/dwc_checklist.Rmd).

# Setup

```{r setup, echo = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Set locale (so we use UTF-8 character encoding):

```{r}
# This works on Mac OS X, might not work on other OS
Sys.setlocale("LC_CTYPE", "en_US.UTF-8")
```

Load libraries:

```{r}
library(tidyverse) # To transform data
library(magrittr)  # For %<>% pipes
library(janitor)   # To clean input data
library(readxl)    # To read Excel files
library(stringr)   # To perform string operations
library(digest)    # To generate hashes
```

Set file paths (all paths should be relative to this script):

```{r}
# Raw files:
raw_data_file = "../data/raw/AI_2016_Boets_etal_Supplement.xls"
sources_file = "../data/raw/sources.tsv"

# Processed files:
dwc_taxon_file = "../data/processed/dwc_checklist/taxon.csv"
dwc_distribution_file = "../data/processed/dwc_checklist/distribution.csv"
dwc_profile_file = "../data/processed/dwc_checklist/speciesprofile.csv"
dwc_description_file = "../data/processed/dwc_checklist/description.csv"
```

# Read data

Create a data frame `raw_data` from the source data:

```{r}
# Read the source data:
raw_data <- read_excel(raw_data_file, sheet = "checklist", na = "NA") 
sources <- read.table(sources_file, sep = "\t", quote = "", colClasses = "character",  fileEncoding = "UTF8", header = T)

# Clean the data somewhat: remove empty rows if present
raw_data %<>%
  remove_empty_rows() %>%     # Remove empty rows
  clean_names()               # Have sensible (lowercase) column names
```

## Generate taxonID

To uniquely identify a taxon in the taxon core and reference taxa in the extensions, we need a `taxonID`. Since we need it in all generated files, we generate it here in the raw data frame. It is a combination of `dataset-shortname:taxon:` and a hash based on the scientific name. As long as the scientific name doesn't change, the ID will be stable:

```{r}
# Vectorize the digest function (The digest() function isn't vectorized. So if you pass in a vector, you get one value for the whole vector rather than a digest for each element of the vector):
vdigest <- Vectorize(digest)

# Generate taxonID:
raw_data %<>% mutate(taxon_id = paste("alien-macroinvertebrates-checklist", "taxon", vdigest(species, algo="md5"), sep = ":"))
```

## Use full references

The raw data contains (abbreviated) citations in `raw_reference` (e.g. `van Haaren and Soors 2009`). We want to use full reference (e.g. `van Haaren T, Soors J (2009) Sinelobus stanfordi (Richardson, 1901): a new crustacean invader in Europe. Aquatic Invasions 4: 703–711, https://doi.org/10.3391/ai.2009.4.4.20`), which are stored in a separate sources file. We will join the two here, so we can use this information in the core and/or extensions.

Clean `raw_reference` somewhat:

```{r}
raw_data %<>% mutate(reference = recode(reference,
  "Adam  and Leloup 1934" = "Adam and Leloup 1934", # Remove whitespace
  "Van  Haaren and Soors 2009" = "van Haaren and Soors 2009", # Remove whitespace and lowercase "van"
  "This study" = "Boets et al. 2016",
  "Nyst 1835; Adam 1947" = "Nyst 1835 | Adam 1947" )
)
```

Left join `raw_data` with `sources` on `references` = `citation`:

```{r}
raw_data %<>% 
  left_join(sources, by = c("reference" = "citation"))
```

Show result (note: `Boets. et al. unpub data` and `Collection RBINS` don't have a full reference):

```{r}
raw_data %>% 
  mutate(full_reference = substr(full_reference, 1, 50)) %>% # Shorten full_reference to make it easier to display
  select(reference, full_reference) %>%
  group_by_all %>%
  summarize(records = n()) %>%
  arrange(reference)
```

Add prefix `raw_` to all column names to avoid name clashes with Darwin Core terms:

```{r}
colnames(raw_data) <- paste0("raw_", colnames(raw_data))
```
 
Preview data:

```{r}
head(raw_data)
```

# Create taxon core

```{r start_taxon}
taxon <- raw_data
```

## Term mapping

Map the data to [Darwin Core Taxon](http://rs.gbif.org/core/dwc_taxon_2015-04-24.xml).
 
### language

```{r}
taxon %<>% mutate(language = "en")
```

### license

```{r}
taxon %<>% mutate(license = "http://creativecommons.org/publicdomain/zero/1.0/")
```

### rightsHolder

```{r}
taxon %<>% mutate(rightsHolder = "Ghent University Aquatic Ecology")
```

### datasetID

```{r}
taxon  %<>% mutate(datasetID = "https://doi.org/10.15468/yxcq07")
```

### datasetName

```{r}
taxon %<>% mutate(datasetName = "Inventory of alien macroinvertebrates in Flanders, Belgium")
```

### taxonID

```{r}
taxon %<>% mutate(taxonID = raw_taxon_id)
```

### scientificName

```{r}
taxon %<>% mutate(scientificName = raw_species) %>% mutate(scientificName = str_trim(scientificName))
```

Verify that `scientificName` contains unique values:

```{r}
any(duplicated(taxon$scientificName)) # Should be FALSE
```

### kingdom

```{r}
taxon %<>% mutate(kingdom = "Animalia")
```

### phylum

```{r}
taxon %<>% mutate(phylum = recode(raw_phylum,
  "Crustacea" = "Arthropoda" # Crustacea is a subphylum, the correct phylum is "Arthropoda"
))
```

### order

```{r}
taxon %<>%
  mutate(order = recode(raw_order,
    "Veneroidea" = "Venerida" # Correct typo
  )) %>%
  mutate(order = str_trim(order)) # Oddly, not all whitespace got detected and trimmed by read_excel
```

### family

```{r}
taxon %<>% mutate(family = raw_family)
```

### taxonRank

```{r}
taxon %<>% mutate(taxonRank = case_when(
  raw_species == "Dreissena rostriformis bugensis" ~ "subspecies",
  raw_species != "Dreissena rostriformis bugensis" ~ "species")
)
```

### nomenclaturalCode

```{r}
taxon %<>% mutate(nomenclaturalCode = "ICZN")
```
 
## Post-processing

Remove the original columns:

```{r}
taxon %<>% select(-starts_with("raw_"))
```

Preview data:

```{r}
head(taxon)
```

Save to CSV:

```{r}
write.csv(taxon, file = dwc_taxon_file, na = "", row.names = FALSE, fileEncoding = "UTF-8")
```

# Create distribution extension

```{r start_distribution}
distribution <- raw_data
```

## Term mapping

Map the data to [Species Distribution](http://rs.gbif.org/extension/gbif/1.0/distribution.xml).

### taxonID

```{r}
distribution %<>% mutate(taxonID = raw_taxon_id)
```

### locationID

```{r}
distribution %<>% mutate(locationID = "ISO_3166-2:BE-VLG")
```

### locality

```{r}
distribution %<>% mutate(locality = "Flemish Region")
```

### countryCode

```{r}
distribution %<>% mutate(countryCode = "BE")
```

### occurrenceStatus

```{r}
distribution %<>% mutate(occurrenceStatus = "present")
```

### establishmentMeans

We use the [GBIF controlled vocabulary](http://rs.gbif.org/vocabulary/gbif/establishment_means.xml) for this term. For this dataset, all species are `introduced` (= alien):

```{r}
distribution %<>% mutate(establishmentMeans = "introduced")
```

### eventDate

The distribution information applies to a certain date range, which we will express here as an ISO 8601 date `yyyy/yyyy` (`start_year/end_year`).

The date information can be found in `raw_first_occurrence_in_flanders`, but it is currently expressed in different formats: `yyyy`, `< yyyy`, `<yyyy`, `before yyyy` and `yyyy-yyyy`:

```{r}
distribution %>%
  distinct(raw_first_occurrence_in_flanders) %>%
  arrange(raw_first_occurrence_in_flanders)
```

We remove the special characters from `raw_first_occurrence_in_flanders` and save as `year`:

```{r}
distribution %<>% mutate(year = str_replace_all(raw_first_occurrence_in_flanders, "(< |before |<)", ""))
```

Then, we split the information on `-` to get a `start_year` and `end_year`. Most records won't have an `end_year`. For those we'll consider the publication year of Boets et al. (2016) as the date when the presence of the species was last verified:

```{r}
distribution %<>%
  separate(year, into = c("start_year", "end_year"), sep = "-") %>% # Separate year
  mutate(end_year = case_when(
    is.na(end_year) ~ "2016",
    TRUE ~ end_year)
  )
```

Create `eventDate` by combining `start_year/end_year`:

```{r}
distribution %<>% mutate(eventDate = paste(start_year, end_year, sep = "/"))
```

Compare formatted dates with original dates in `raw_first_occurrence_in_flanders`:

```{r}
distribution %>%
  distinct(raw_first_occurrence_in_flanders, eventDate) %>%
  arrange(raw_first_occurrence_in_flanders)
```

### source

```{r}
distribution %<>% mutate(source = raw_full_reference)
```

## Post-processing

Remove the original columns:

```{r}
distribution %<>% select(-starts_with("raw_"), -start_year, -end_year)
```

Preview data:

```{r}
distribution %>%
  mutate(source = substr(source, 1, 10)) %>% # Shorten source to make it easier to display
  head()
```

Save to CSV:

```{r}
write.csv(distribution, file = dwc_distribution_file, na = "", row.names = FALSE, fileEncoding = "UTF-8")
```

# Create species profile extension

In this extension will express broad habitat characteristics (e.g. `isMarine`) of the species.

```{r start_species_profile}
species_profile <- raw_data
```

Habitat information can be found in `raw_salinity_zone`, which describes whether a species is found in brackish (`B`), freshwater (`F`), marine (`M`) or combined (`B/M` or `F/B`) salinity zones.

Show unique values:

```{r}
species_profile %>%
  distinct(raw_salinity_zone) %>%
  arrange(raw_salinity_zone)
```

This is a manually created overview of how we will map the `raw_salinity_zone` to `isMarine`, `isFreshwater` and `isTerrestrial`. Note: the latter term is added for completeness, but will be `FALSE` for all species in this dataset:

```{r, echo = FALSE}
as.data.frame(matrix(data = c(
  "F", "FALSE", "TRUE", "FALSE",
  "M", "TRUE", "FALSE", "FALSE",
  "B/M", "TRUE", "FALSE", "FALSE",
  "F/B", "FALSE", "TRUE", "FALSE",
  "B", "TRUE", "TRUE", "FALSE"),
  nrow = 5, ncol = 4, byrow = TRUE,
  dimnames = list (c(1:5), c("salinity zone", "isMarine", "isFreshwater", "isTerrestrial"))
))
```

## Term mapping

Map the data to [Species Profile](http://rs.gbif.org/extension/gbif/1.0/speciesprofile.xml).

### taxonID

```{r}
species_profile %<>% mutate(taxonID = raw_taxon_id)
```

### isMarine

```{r}
species_profile %<>% mutate(isMarine = case_when(
  raw_salinity_zone == "M" | raw_salinity_zone == "B/M" | raw_salinity_zone == "B" ~ "TRUE",
  TRUE ~ "FALSE"
))
```

### isFreshwater

```{r}
species_profile %<>% mutate(isFreshwater = case_when(
  raw_salinity_zone == "F" | raw_salinity_zone == "F/B" | raw_salinity_zone == "B" ~ "TRUE",
  TRUE ~"FALSE"
))
```

### isTerrestrial

```{r}
species_profile %<>% mutate(isTerrestrial = "FALSE")
```

Show mapped values:

```{r}
species_profile %>%
  select(raw_salinity_zone, isMarine, isFreshwater, isTerrestrial) %>%
  group_by_all() %>%
  summarize(records = n())
```

## Post-processing

Remove the original columns:

```{r}
species_profile %<>% select(-starts_with("raw_"))
```

Preview data:

```{r}
species_profile %>% head()
```

Save to CSV:

```{r}
write.csv(species_profile, file = dwc_profile_file, na = "", row.names = FALSE, fileEncoding = "UTF-8")
```

# Create description extension

In the description extension we want to include several important characteristics about the species.
 
## Native range

Native range information (e.g. `South-America`) can be found in `raw_origin`. We create a separate dataframe to process this information:

```{r start_native_range}
native_range <- raw_data
```

Show unique values:

```{r}
native_range %>%
  distinct(raw_origin) %>%
  arrange(raw_origin)
```

`raw_origin` contains multiple values (currently not more than 3), so we separate it in 3 columns:

```{r}
native_range %<>% separate(raw_origin, 
  into = c("native_range_1", "native_range_2", "native_range_3"),
  sep = ", ",
  remove = TRUE,
  convert = FALSE,
  extra = "merge",
  fill = "right"
)
```

Gather in a `key` and `value` column:

```{r}
native_range %<>% gather(
  key, value,
  native_range_1, native_range_2, native_range_3,
  na.rm = TRUE, # Also removes records for which there is no native_range_1
  convert = FALSE
)
```

Map values:

```{r}
native_range %<>% mutate(mapped_value = recode(value,
  "East-Asia" = "East Asia",
  "East-Europe" = "Eastern Europe",
  "Indio-Pacific" = "Indo-Pacific",
  "North-Africa" = "North Africa",
  "North-America" = "Northern America",
  "Northeast-Asia" = "Northeast Asia",
  "South-America" = "South America",
  "South-Europe" = "Southern Europe",
  "Southeast-Asia" = "Southeast Asia",
  "Southern hemisphere" = "Southern Hemisphere",
  "USA" = "United States",
  "West-Africa" = "West Africa",
  "West-Atlantic" = "Western Atlantic"
))
```

Show mapped values:

```{r}
native_range %>%
  select(value, mapped_value) %>%
  group_by(value, mapped_value) %>%
  summarize(records = n()) %>%
  arrange(value)
```

Drop the `key` and `value` columns and rename `mapped_value` as `description`:

```{r}
native_range %<>%
  select(-key, -value) %>%
  rename(description = mapped_value)
```

Create a `type` field to indicate the type of description:

```{r}
native_range %<>% mutate(type = "native range")
```

## Pathway of introduction

Pathway information (e.g. `aquaculture`) can be found in `raw_pathway_of_introduction`. This original information was interpreted by @timadriaens to a more standardized value in `raw_pathway_mapping` (with remarks about this transformation in `raw_pathway_mapping_remarks`). We create a separate dataframe to process this information:

```{r start_pathway}
pathway <- raw_data
```

Show unique values:

```{r}
native_range %>%
  distinct(raw_pathway_mapping) %>%
  arrange(raw_pathway_mapping)
```

`raw_pathway_mapping` contains multiple values (currently not more than 3), so we separate it in 3 columns:

```{r}
pathway %<>% separate(raw_pathway_mapping, 
  into = c("pathway_1", "pathway_2", "pathway_3"),
  sep = " \\| ",
  remove = TRUE,
  convert = FALSE,
  extra = "merge",
  fill = "right"
)
```

Gather in a `key` and `value` column:

```{r}
pathway %<>% gather(
  key, value,
  pathway_1, pathway_2, pathway_3,
  na.rm = TRUE, # Also removes records for which there is no pathway_1
  convert = FALSE
)
```

We use the [CBD 2014 pathway vocabulary](https://www.cbd.int/doc/meetings/sbstta/sbstta-18/official/sbstta-18-09-add1-en.pdf) to standardize this information. The vocubulary has [these values](https://github.com/trias-project/vocab/tree/master/vocabulary/pathway).

Map values:

```{r}
pathway %<>% mutate (mapped_value = recode(value,
  "Aquaculture" = "escape_aquaculture",
  "Aquaculture / mariculture" = "escape_aquaculture",
  "Contaminant on animals (except parasites, species transported by host/vector)" = "contaminant_on_animals",
  "Interconnected waterways/basins/seas" = "corridor_water",
  "Mariculture" = "escape_aquaculture",
  "Other means of transport" = "stowaway_other",
  "Pet/aquarium/terrarium species (including live food for such species )" = "escape_pet",
  "Ship/boat ballast water" = "stowaway_ballast_water",
  "Ship/boat hull fouling" = "stowaway_hull_fouling"))
```

Add the prefix `cbd_2014_pathway:` to refer to this standard:

```{r}
pathway %<>% mutate(mapped_value = paste ("cbd_2014_pathway", mapped_value, sep = ":"))
```

Show mapped values:

```{r}
pathway %>%
  select(value, mapped_value) %>%
  group_by(value, mapped_value) %>%
  summarize(records = n()) %>%
  arrange(value)
```

Drop the `key` and `value` columns and rename `mapped_value` as `description`:

```{r}
pathway %<>%
  select(-key, -value) %>%
  rename(description = mapped_value)
```

Create a `type` field to indicate the type of description:

```{r}
pathway %<>% mutate(type = "pathway")
```

### Invasion stage

There is no invasion stage information (e.g. `casual`) in the raw data, but we want to add it none the less. We create a separate dataframe to process this information:

```{r start_invasion_stage}
invasion_stage <- raw_data
```

We use the [invasion stage vocabulary from Blackburn et al. (2011)](http://doc.rero.ch/record/24725/files/bach_puf.pdf) to standardize this information.

Here, we consider all species to be `established` as they come from live samples in running waters. We decided **not** to use the terms `naturalized` (because often, there's no sensible criterium to distinguish between casual/naturalized of naturalized/established) and `invasive` (which is a term that can only be applied after a risk assessment).

```{r}
invasion_stage %<>% mutate(description = "established")
```

Create a `type` field to indicate the type of description:

```{r}
invasion_stage %<>% mutate(type = "invasion stage")
```

## Union descriptions

Union native range, pathway of introduction and invasion stage:

```{r start_description_ext}
description_ext <- bind_rows(native_range, pathway, invasion_stage)
```

## Term mapping
 
Map the data to [Taxon Description](http://rs.gbif.org/extension/gbif/1.0/description.xml).

### taxonID

```{r}
description_ext %<>% mutate(taxonID = raw_taxon_id)
```

### description

```{r}
description_ext %<>% mutate(description = description)
```

### type

```{r}
description_ext %<>% mutate(type = type)
```

### source

```{r}
description_ext %<>% mutate(source = raw_full_reference)
```

### language

```{r}
description_ext %<>% mutate(language = "en")
```

## Post-processing

Remove the original columns:

```{r}
description_ext %<>% select(-starts_with("raw_"))
```

Move `taxonID` to the first position:

```{r}
description_ext %<>% select(taxonID, everything())
```

Sort on `taxonID`:

```{r}
description_ext %<>% arrange(taxonID)
```

Preview data:

```{r}
description_ext %>% 
  mutate(source = substr(source, 1, 10)) %>% # Shorten source to make it easier to display
  head()
```

Save to CSV:

```{r}
write.csv(description_ext, file = dwc_description_file, na = "", row.names = FALSE, fileEncoding = "UTF-8")
```
